---
title: "ADS506 Assignment 1.2 <Rain in Australia Revised>"
author: "Robert Marriott"
date: "`31 October 2024`"
output: pdf_document
---

https://www.kaggle.com/datasets/jsphyg/weather-dataset-rattle-package

**Submission Requirements**:

-   **Dataset Description**: Provide a brief description of your dataset, its source, and why you chose it.

-   **Time Series Plot:** Create a reproducible time series plot with an appropriate title and axes labels (include code that shows the import, transformation and plotting).

-   **Preliminary Observations**: Share any initial patterns or trends you observe in the data. This doesn't have to be detailed, just your first impressions.

-   **Relevance**: Explain the potential value in forecasting this data, either in a business context, research scenario, or other relevance.

\newpage

```{r warning=FALSE, message=FALSE}
library(ggplot2)
library(readr)
library(dplyr)
```

## Data Source

Include public links to data if it is too large to upload (do not upload datasets larger than 50MB).

## Importing the Data

```{r}
weather_data <- read.csv("/Users/bobbymarriott/Desktop/ADS-506/weatherAUS.csv")
```

## Time Series Plot

```{r}
# Convert Date to Date type and remove rows with NA in Date or Rainfall
weather_data$Date <- as.Date(weather_data$Date, format="%Y-%m-%d")
weather_data <- weather_data %>% filter(!is.na(Date) & !is.na(Rainfall))

# Aggregate data to get monthly rainfall totals
monthly_rainfall <- weather_data %>%
  group_by(Location, Month = as.Date(format(Date, "%Y-%m-01"))) %>%
  summarize(MonthlyRainfall = sum(Rainfall, na.rm = TRUE))

# Filter for Sydney
monthly_rainfall <- monthly_rainfall %>% filter(Location == "Sydney")

ggplot(monthly_rainfall, aes(x = Month, y = MonthlyRainfall)) +
  geom_line(color = "blue", size = 0.5) +
  labs(title = "Monthly Rainfall in Sydney",
       x = "Date",
       y = "Total Monthly Rainfall (mm)") +
  theme_minimal()
```

## Discussion

This dataset provides monthly aggregated rainfall data collected over 10 years from weather stations across Australia. Key variables include total monthly rainfall, minimum temperature, and the target variable RainTomorrow, making it well-suited for examining long-term rainfall trends. The graph above displays monthly rainfall for Sydney, revealing high variability with notable peaks representing periods of more rainfall but no obvious seasonal pattern.  

Monthly rainfall trends offer valuable insights for sectors like agriculture and water management, where understanding variability in rainfall impact how industries plan ahead. By focusing on monthly patterns, we can better predict significant changes in rainfall, aiding industries in preventing issues during droughts or damaages from heavy rainfall.  